---
title: "HW2: Exploring Mass Shootings in America"
author: "Julia Du"
output: github_document
---

## Get the data

```{r, echo = TRUE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)        # load functions for formatting tables
library(ggthemes) 

# get data from rcfss package
# install latest version if not already installed
# devtools::install_github("uc-cfss/rcfss")
library(rcfss)

# load the data
data("mass_shootings")
mass_shootings
```

## Generate a data frame that summarizes the number of mass shootings per year. Print the data frame as a formatted `kable()` table.

```{r}
mass_shootings %>%
  group_by(year) %>%
  summarize(
    count = n()
  ) %>%
  kable(caption = "Mass shootings (1982-2019)", col.names = c("Year", "Number of mass shootings"))
```

## Generate a bar chart that identifies the number of mass shooters associated with each race category. The bars should be sorted from highest to lowest.

```{r}
library(forcats)
require(forcats)

mass_shootings %>%
  drop_na(race) %>%
  group_by(race) %>%
  summarize(
    count = n() 
  ) %>% ungroup() %>%
  ggplot(mapping = aes(x=fct_reorder(race, -count), y = count)) + #we use -count to sort by descending
  geom_col() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Number of incidents") +
  theme_solarized()
```

## Generate a boxplot visualizing the number of total victims, by type of location. Redraw the same plot, but remove the Las Vegas Strip massacre from the dataset.

```{r}
# w/ the LV Strip massacre
ggplot(data = mass_shootings, mapping = aes(x = location_type, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Type of location of incident", y = "Total number of victims") +
  theme_solarized()

#omitting LV Strip massacre
mass_shootings %>%
    filter(case != "Las Vegas Strip massacre") %>%
  ggplot(mapping = aes(x = location_type, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", subtitle = "Omitting the Las Vegas Strip massacre (2017)", x = "Type of location of incident", y = "Total number of victims") +
  theme_solarized()
```

## How many white males with prior signs of mental illness initiated a mass shooting after 2000?

```{r}
#filter by race = "White", male = TRUE, prior_mental_illness = "Yes", year > 2000
mass_shootings %>%
  filter(race == "White", male, prior_mental_illness == "Yes", year > 2000) %>% 
  mutate(white_after2000 = row_number()) %>%
  glimpse()
```
*20 white males with prior signs of mental illness initiated a mass shooting after 2000.*

## Which month of the year has the most mass shootings? Generate a bar chart sorted in chronological order to provide evidence of your answer.

```{r}
mass_shootings %>%
#  drop_na(month) %>% turns out we didn't need to do NA here since there are no NA values for months
  mutate(month = factor(month,
                  levels = month.abb)) %>%
  count(month) %>%
  ggplot(mapping = aes(x=month, y = n)) +
  geom_col() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Month of incident", y = "Number of incidents") +
  scale_y_continuous(breaks=seq(0,12,2)) +
  theme_solarized()
```

*February and June have the most mass shootings.*

## How does the distribution of mass shooting fatalities differ between white and black shooters? What about white and latino shooters?

```{r}
#need to group shooters by race. filter out only white, black, or latino. 
mass_shootings %>%
  group_by(race) %>%
  filter(
    race %in% c("White", "Black", "Latino")
  ) %>%
  ggplot(mapping = aes(x = fatalities)) +
  geom_histogram() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Number of fatalities per incident", y = "Number of incidents") +
  facet_grid(race ~ ., scales = "free") +
  theme_solarized()


mass_shootings %>%
  group_by(race) %>%
  filter(
    race %in% c("White", "Black", "Latino")
  ) %>%
  ggplot(mapping = aes(x = fatalities, color = race)) +
  geom_freqpoly() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Number of fatalities per incident", y = "Number of incidents") +
  scale_color_discrete(name="Race") +
  theme_clean()
```

## Are mass shootings with shooters suffering from mental illness different from mass shootings with no signs of mental illness in the shooter? Assess the relationship between mental illness and total victims, mental illness and race, and the intersection of all three variables.

```{r}
#need to create graphs between mental illness & total victims. mental illness & race. then create a final graph between mental illness, race, & total victims.

#mental illness vs total victims
mass_shootings %>%
  drop_na(prior_mental_illness) %>%
  ggplot(mapping = aes(x = prior_mental_illness, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Evidence of prior mental illness from perpetrator", y = "Total number of victims") +
  theme_solarized()

##mental illness vs race 
ggplot(data = na.omit(mass_shootings[, c("prior_mental_illness", "race")]), mapping = aes(x = prior_mental_illness, fill = race)) +
  geom_bar(position = position_dodge2(width = 0.9, preserve = "single")) +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Evidence of prior mental illness from perpetrator", y = "Number of perpetrators") +
  scale_fill_discrete(name="Race of perpetrator") +
  theme_clean()

#victims vs race
mass_shootings %>%
  drop_na(race) %>%
  ggplot(mapping = aes(x = race, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Total number of victims") +
  theme_solarized()

#victims vs race - w/ log y-axis
mass_shootings %>%
  drop_na(race) %>%
  ggplot(mapping = aes(x = race, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Total number of victims") +
    scale_y_continuous(trans = "log10") +
  theme_solarized()

#victims vs race, w/ LV massacre omitted
mass_shootings %>%
  filter(case != "Las Vegas Strip massacre") %>%
  drop_na(race) %>%
  ggplot(mapping = aes(x = race, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", subtitle = "Omitting the Las Vegas Strip massacre (2017)", x = "Race of perpetrator", y = "Total number of victims") +
  theme_solarized()

#victims vs race, by mental illness 
mass_shootings %>%
  drop_na(race, prior_mental_illness) %>%
  ggplot(mapping = aes(x = total_victims, y = race, fill = prior_mental_illness)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Total number of victims", y = "Race of perpetrator") +
  scale_fill_discrete(name="Evidence of prior mental illness from perpetrator") +
  theme_clean() +
  theme(legend.position="bottom") 

#mental illness vs victims, by race 
mass_shootings %>%
  drop_na(race, prior_mental_illness) %>%
  ggplot(mapping = aes(x = prior_mental_illness, y = total_victims, fill = race)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Evidence of prior mental illness from perpetrator", y = "Total number of victims") +
  scale_fill_discrete(name="Race of perpetrator") +
  theme_clean() +
  theme(legend.position="bottom")
```


*In general, it doesn't seem like evidence of prior mental illness has that large of an effect on the number of victims. The median # of total victims for perpetrators with mental illness is a bit higher, but it's relatively close to perpetrators without mental illness. However, perpetrators with mental illness do have a much larger upper bound of total victims, especially for white perpetrators.*
White perpetrators seem to have a higher total #s of victims, with the exception of perpetrators of race "Other." Perpetrators with "Other" race have a noticeably higher median total # of victims than other races, but because there are only 5 such perpetrators (i.e.the sample size is much smaller), we should be careful about concluding anything from that. Moreover, there was no info on if any of these "Other" race perpetrators showed evidence of prior mental illness.
It seems that white perpetrators with mental illness tended to have higher total # of victims. For races that had sufficient prior mental illness info available, it seems like perpetrators with evidence of prior mental illness had slightly higher victim totals, regardless of race.
Across most races, it seems that it was more common for perpetrators to have evidence of prior mental illness than not. 

## Session info

```{r, echo = TRUE}
devtools::session_info()
```


