---
title: "HW2: Exploring Mass Shootings in America"
author: "Julia Du"
output: github_document
---

## Get the data

```{r, echo = TRUE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)        # load functions for formatting tables

# get data from rcfss package
# install latest version if not already installed
# devtools::install_github("uc-cfss/rcfss")
library(rcfss)

# load the data
data("mass_shootings")
mass_shootings
```

## Generate a data frame that summarizes the number of mass shootings per year. Print the data frame as a formatted `kable()` table.

```{r}
#shootings_yr = group_by(.data = mass_shootings, year)
#summarize(.data = shootings_yr, count=n())

### same as above, but with a pipe
shootings_yrpipe = mass_shootings %>%
  group_by(year) %>%
  summarize(
    count = n()
  )
kable(shootings_yrpipe, caption = "Mass shootings (1982-2019)", col.names = c("Year", "Number of mass shootings"))
```

## Generate a bar chart that identifies the number of mass shooters associated with each race category. The bars should be sorted from highest to lowest.

```{r}
### this is unnecessary. ask TAs about how i'd make a graph from this table
mass_shootings %>%
  drop_na(race) %>%
  group_by(race) %>%
  summarize(
    count = n() 
  ) %>% ungroup() %>%
  ggplot(mapping = aes(x=fct_reorder(race, -count), y = count)) +
  geom_col()

library(forcats)
require(forcats)

ggplot(data = mass_shootings[!is.na(mass_shootings$race), ], mapping = aes(x = fct_infreq(race))) +
 geom_bar() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Number of incidents")

####ask TAs about why the square brackets work, i.e. formatting of [!is.na(df$column),]
#above works. can also do
#ggplot(data = na.omit(mass_shootings[, c("race")]), aes(x=fct_infreq(race))) + 
 # geom_bar() +
  #labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Number of incidents")
#%>%
  arrange(desc(count)) %>%
  na.omit()

```

## Generate a boxplot visualizing the number of total victims, by type of location. Redraw the same plot, but remove the Las Vegas Strip massacre from the dataset.

```{r}
# w/ the LV Strip massacre
ggplot(data = mass_shootings, mapping = aes(x = location_type, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Type of location of incident", y = "Total number of victims")

#omitting LV Strip massacre
noLV = mass_shootings %>%
    filter(case != "Las Vegas Strip massacre") 

ggplot(data = noLV, mapping = aes(x = location_type, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", subtitle = "Omitting the Las Vegas Strip massacre (2017)", x = "Type of location of incident", y = "Total number of victims")
```

## How many white males with prior signs of mental illness initiated a mass shooting after 2000?

```{r}
#filter by race = "White", male = TRUE, prior_mental_illness = "Yes", year > 2000
mass_shootings %>%
  filter(race == "White", male == TRUE, prior_mental_illness == "Yes", year > 2000) %>% 
  mutate(white_after2000 = row_number()) %>%
  glimpse()

#### ask TAs about if you ever need to put TRUE or FAlSE in quotation marks. 
#for this case, can just write "male". if looking for non-male, do "!male"

#### also quotations vs backticks
#backticks when you select columns
#e.g. filter(`Race of Shooter` == “White”)

```
**20 white males with prior signs of mental illness initiated a mass shooting after 2000.**

## Which month of the year has the most mass shootings? Generate a bar chart sorted in chronological order to provide evidence of your answer.

```{r}
#for table, need to group shootings by month. then arrange to see which month is biggest. 
#### this 1st block is unnecessary for answering the q!
mass_shootings %>%
  mutate(month = factor(month,
                  levels = month.abb)) %>%
  group_by(month) %>%
  summarize(count = n() )   %>%
  ungroup() %>%
  ggplot(mapping = aes(x=month, y = count)) +
  geom_col()
  
mass_shootings %>%
  mutate(month = factor(month,
                  levels = month.abb)) %>%
  count(month) %>%
  ggplot(mapping = aes(x=month, y = n)) +
  geom_col()

month_shootings = mass_shootings
month_shootings$month = factor(mass_shootings$month,
                  levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
#levels = month.abb

ggplot(data = month_shootings[!is.na(month_shootings$month), ], mapping = aes(x = month)) +
 geom_bar() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Month of incident", y = "Number of incidents") +
  scale_y_continuous(breaks=c(0,2,4,6,8,10,12))
```
**February and June have the most mass shootings.**

## How does the distribution of mass shooting fatalities differ between white and black shooters? What about white and latino shooters?

```{r}
#need to group shooters by race. filter out everything not white, black, or latino. 
#must sum the # of fatalities for each race
#then draw graphs to look at fatalities (bar graph & line graph)
fatal_compare = mass_shootings %>%
  group_by(race) %>%
  summarize(
    deaths = sum(fatalities, na.rm = TRUE)
  ) %>%
  filter(
    race %in% c("White", "Black", "Latino")
  ) 
### DON'T THINK i need the above. gives a table of total fatalities by shooter race, but looking for distr (Not #s)


ggplot(data = subset(mass_shootings, 
                     race %in% c("White", "Black", "Latino")),
mapping = aes(x = fatalities)) +
  geom_histogram() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Number of fatalities per incident", y = "Number of incidents") + 
  facet_grid(race ~ ., scales = "free")
### DO I NEED TO CHANGE THE X-AXIS OF THE FACET GRID? THINK IT'S EASIER TO COMPARE W/ SAME X-AXIS

#add onto current pipe, can use facet_wrap
#to make graphs pretty, use ggthemes

#attempt at smooth graph
ggplot(data = subset(mass_shootings, 
                     race %in% c("White", "Black", "Latino")),
mapping = aes(x = fatalities, color = race)) +
  geom_freqpoly() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Number of fatalities per incident", y = "Number of incidents") +
  scale_color_discrete(name="Race")

```

## Are mass shootings with shooters suffering from mental illness different from mass shootings with no signs of mental illness in the shooter? Assess the relationship between mental illness and total victims, mental illness and race, and the intersection of all three variables.

```{r}
#need to create graphs between mental illness & total victims. mental illness & race. 
#then create a final graph between mental illness, race, & total victims. ~~~use color = mental illness 

#mental illness vs total victims
ggplot(data = mass_shootings[!is.na(mass_shootings$prior_mental_illness), ], mapping = aes(x = prior_mental_illness, y = total_victims)) +
 geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Evidence of prior mental illness from perpetrator", y = "Total number of victims")

### other way to do mental ill vs victims - gives same graph!
ggplot(data = na.omit(mass_shootings[, c("prior_mental_illness", "total_victims")]), mapping = aes(x = prior_mental_illness, y = total_victims)) + 
  geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Evidence of prior mental illness from perpetrator", y = "Total number of victims")

##mental illness vs race - weird graph, ask TA!!!!!
#ggplot(data = na.omit(mass_shootings[, c("prior_mental_illness", "race")]), mapping = aes(x = prior_mental_illness, y = race)) + 
 # geom_boxplot() +
  #labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Evidence of prior mental illness from perpetrator", y = "Race of perpetrator")

#victims vs race
ggplot(data = na.omit(mass_shootings[, c("race", "total_victims")]), mapping = aes(x = race, y = total_victims)) + 
  geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Total number of victims")
#victims vs race, w/ LV massacre omitted
ggplot(data = na.omit(noLV[, c("race", "total_victims")]), mapping = aes(x = race, y = total_victims)) + 
  geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Race of perpetrator", y = "Total number of victims")

#victims vs race, by mental illness
ggplot(data = na.omit(mass_shootings[, c("race", "total_victims","prior_mental_illness")]), mapping = aes(x = total_victims, y = race, fill = prior_mental_illness)) + 
  geom_boxplot() +
  labs(title = "Mass shootings in the U.S. (1982-2019)", x = "Total number of victims", y = "Race of perpetrator") +
  scale_fill_discrete(name="Evidence of prior mental illness from perpetrator") +
  theme(legend.position="bottom")

```
**comments on graph - CHECK THIS**
*In general, it doesn't seem like evidence of prior mental illness has that large of an effect on the number of victims. The median # of total victims for perpetrators with mental illness is a bit higher, but it's relatively close to perpetrators without mental illness. However, perpetrators with mental illness do have a much larger upper bound of total victims.*
*White perpetrators seem to have higher total #s of victims, with the exception of perpetrators with their race being categorized as "other." Perpetrators with "Other" race likely have an extremely high median total # of victims because there are only 5 such perpatrators, so the sample size is much smaller.
*seems white perpatrators with mental illness tend to have higher total # of victims. For races that had mental illness info available, it seems like perpetrators w/ mental illness had higher victim totals, regardless of race.

## Session info

```{r, echo = TRUE}
devtools::session_info()
```


